---
title: "stretch02"
subtitle:"Predicting Civil Servant Exit"
author: "Benjamin Reese"
format: pdf
---

## Setup

```{r packages and data loading}
## Packages
library(tidyverse)
library(tidymodels)
library(lubridate)
library(readr)

## Data Loading
sep_data <- read.csv("SEPDATA_FY2005-2009.TXT")
load("~/Fall 2022/Data Science/stretch02/data/fed.RData")
app_data <- read_csv("data/EAD+2.0+quarter+101019.csv")
```

### (1)

I am loading and joining three datasets together. The main dataset, that includes the variable I want to predict is `sep_data`. This dataset includes information on nearly every federal employee, covering most non-intelligence community agencies, who left federal service from 2005-2009. The goal of this analysis is to predict when federal employees will quit the federal workforce versus leaving for other reasons

```{r cleaning and joining}
## Loading in data on when federal employees quit, creating education and quit variables
sep_data <- sep_data %>%
  mutate(quarter = str_replace_all(EFDATE, "(.{4})(.*)", "\\1-\\2"),
         quarter = quarter(ym(quarter), type = "year.quarter"),
         separation = if_else(SEP == "SC", 1, 0),
         EDLVL = as.numeric(EDLVL)) %>%
  filter(!is.na(EDLVL) & EDLVL != "**")%>%
  mutate(education = case_when(
           EDLVL < 4 ~ "Less Than High School",
           EDLVL == 4 ~ "High School or GED",
           EDLVL == 5 | EDLVL == 6 | EDLVL == 7 | EDLVL == 8 | EDLVL == 9 |
           EDLVL == 10 | EDLVL == 11 | EDLVL == 12 ~ "Less than Bachelors",
           EDLVL == 13 ~ "Bachelors",
           EDLVL > 13 ~ "Graduate")) %>%
  select(separation, education, quarter, SALARY, LOS)

## dropping unused variables, creating a unified presidents and party variable, and filtering to only quarters
## included in the civil servant exit data
econ_data <- dta %>%
  select(-Democrat, -pres1, -pres2, -pres3, -pres4, -pres5, -pres6, -pres7, 
         -pres8, -pres9, -pres10, -pres11, -pres12, -pres13, -pres14, -pres15) %>%
  mutate(admin = case_when(
    president == 1 ~ "Obama",
    president == 2 | president == 3 ~ "Eisenhower",
    president == 4 ~ "H.W. Bush",
    president == 5 | president == 6 ~ "W. Bush",
    president == 7 ~ "Carter",
    president == 8 ~ "Kennedy-Johnson",
    president == 9 ~ "Johnson",
    president == 10 ~ "Nixon-Ford",
    president == 11 ~ "Nixon",
    president == 12 | president == 13 ~ "Reagan",
    president == 14 | president == 15 ~ "Clinton"),
  party = case_when(
    president_party == 2 ~ "R",
    president_party == 1 ~ "D"),
  quarter = quarter(DATE,type = "year.quarter")) %>%
  filter(quarter > 1973.2) %>%
    select(-president, -president_party, -DATE)

## filtering to only include the US, dropping unused variables and renaming
app_data <- app_data %>%
  filter(Country == "United States") %>%
  select(qtr, Approval_Smoothed) %>%
  mutate(approval = Approval_Smoothed,
    quarter = as.numeric(str_replace_all(qtr, 'q', "."))) %>%
  select(-Approval_Smoothed, -qtr)

## joining the three datasets all together into one unified tibble
sep <- sep_data %>%
  left_join(econ_data, by="quarter") %>%
  left_join(app_data, by = "quarter") %>%
  janitor::clean_names()
```

