---
title: "stretch02"
subtitle:"Predicting Civil Servant Exit"
author: "Benjamin Reese"
format: pdf
---

## Setup

```{r packages and data loading}
## Packages
library(tidyverse)
library(tidymodels)
library(lubridate)
library(readr)
library(patchwork)

## Data Loading
sep_data <- read.csv("SEPDATA_FY2005-2009.TXT")
load("~/Fall 2022/Data Science/stretch02/data/fed.RData")
app_data <- read_csv("data/EAD+2.0+quarter+101019.csv")
```

### (a)

I am loading and joining three datasets together. The main dataset, that includes the variable I want to predict is `sep_data`. This dataset includes information on nearly every federal employee, covering most non-intelligence community agencies, who left federal service from 2005-2009. The goal of this analysis is to predict when federal employees will quit the federal workforce versus leaving for other reasons. The dependent variable is coded as quit or not quit. The other two datasets are for factors beyond what OPM tracks that mya contribute to people quitting the federal workforce. These datasets include economic variables like inflation and a variable for presidential approval. The end result of the analysis would be a model that predicts when an employee leaves, that they quit.

```{r cleaning and joining, warning=FALSE,message=FALSE}
## Loading in data on when federal employees quit, creating education and quit variables
sep_data <- sep_data %>%
  mutate(quarter = str_replace_all(EFDATE, "(.{4})(.*)", "\\1-\\2"),
         quarter = quarter(ym(quarter), type = "year.quarter"),
         separation = if_else(SEP == "SC", "Quit", "Not Quit"),
         EDLVL = as.numeric(EDLVL)) %>%
  filter(!is.na(EDLVL) & EDLVL != "**")%>%
  mutate(education = case_when(
           EDLVL < 4 ~ "Less Than High School",
           EDLVL == 4 ~ "High School or GED",
           EDLVL == 5 | EDLVL == 6 | EDLVL == 7 | EDLVL == 8 | EDLVL == 9 |
           EDLVL == 10 | EDLVL == 11 | EDLVL == 12 ~ "Less than Bachelors",
           EDLVL == 13 ~ "Bachelors",
           EDLVL > 13 ~ "Graduate")) %>%
  select(separation, education, quarter, SALARY, LOS)

## dropping unused variables, creating a unified presidents and party variable, and filtering to only quarters
## included in the civil servant exit data
econ_data <- dta %>%
  select(-Democrat, -pres1, -pres2, -pres3, -pres4, -pres5, -pres6, -pres7, 
         -pres8, -pres9, -pres10, -pres11, -pres12, -pres13, -pres14, -pres15) %>%
  mutate(admin = case_when(
    president == 1 ~ "Obama",
    president == 2 | president == 3 ~ "Eisenhower",
    president == 4 ~ "H.W. Bush",
    president == 5 | president == 6 ~ "W. Bush",
    president == 7 ~ "Carter",
    president == 8 ~ "Kennedy-Johnson",
    president == 9 ~ "Johnson",
    president == 10 ~ "Nixon-Ford",
    president == 11 ~ "Nixon",
    president == 12 | president == 13 ~ "Reagan",
    president == 14 | president == 15 ~ "Clinton"),
  party = case_when(
    president_party == 2 ~ "R",
    president_party == 1 ~ "D"),
  quarter = quarter(DATE,type = "year.quarter")) %>%
    select(-president, -president_party, -DATE)

## filtering to only include the US, dropping unused variables and renaming
app_data <- app_data %>%
  filter(Country == "United States") %>%
  select(qtr, Approval_Smoothed) %>%
  mutate(approval = Approval_Smoothed,
    quarter = as.numeric(str_replace_all(qtr, 'q', "."))) %>%
  select(-Approval_Smoothed, -qtr)

## joining the three datasets all together into one unified tibble
sep <- sep_data %>%
  left_join(econ_data, by="quarter") %>%
  left_join(app_data, by = "quarter") %>%
  janitor::clean_names()
```

### (b)

```{r splitting}
## Setting Seed
set.seed(20201124)

## Splitting the Sample
split <- initial_split(sep, prop = 0.7)

## Training and Testing
sep_train <- training(split)
sep_test <- testing(split)
```

### (c)

```{r exploratory data analysis}
## Exploratory Data Analysis
sep_train %>%
  filter(separation == "Quit") %>%
  group_by(quarter) %>%
  count(separation=="Quit") %>%
  ggplot(aes(x=quarter, y=n)) +
  geom_point(col="blue") +
  geom_line(lwd=1, col="light blue") +
  theme_minimal() +
  labs(title = "When Civil Servants Exit", 
       subtitle = "Federal Employees Quit Most in Third Quarters",
       x="Fiscal Year Quarter", y="Number of Employees Quitting",
       caption = "Data Source: Office of Personnel Management")
m1 <- sep_train %>%
  mutate(separation = if_else(separation == "Quit", 1, 0)) %>%
  group_by(quarter) %>%
  summarise(percent_quit = mean(separation)) %>%
  mutate(percent_quit=percent_quit*100) %>%
  ggplot(aes(x=quarter, y=percent_quit)) +
  geom_point(col="blue") +
  geom_line(lwd=1, col="light blue") +
  theme_minimal() +
  lims(y=c(33,43))+
  labs(x="Fiscal Year Quarter", y="Percent of Leavers Who Quit")
m2 / m1

```

